Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-08 main.tex
explication algo best-search & debut de preuve

\noindent
The weight associated by $A$ to $s \in \Delta^*$
is defined according to empty stack semantics:
%
\begin{equation}
A(s)  =
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\weight_A\bigl(\config{q}{\bot}, s, \config{q'}{\bot}\bigr)
\mathop{\otimes} \mathsf{out}(q').
\label{eq:SWVPA-value}
\end{equation}

@@

%\subsection{Best Search}
\label{sec:best}\label{sec:search}
\noindent
We shall now present a procedure for searching, for a~$\SWVPA$ $A$,
a word %$w \in \Delta^*$
of minimal weight for~$A$,
as stated in the following proposition. %\wrt~$\leq_\oplus$,
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a Dijkstra algorithm computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
it holds that:
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
Since $\Semiring$ is complete, the infinite sum in \eqref{eq:bbot} is well defined,
and,  it is the minimum in $\Delta^*$,
\wrt~$\leq_\oplus$, % this ordering.
of the fonction
$s \mapsto \weight_A(\config{q}{\sigma}, s, \config{q'}{\sigma})$.
%
Hence, following~\eqref{eq:SWVPA-value}, and the associativity and commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is
\(
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\).

\noindent
In order to compute the above function $b_\bot : Q \times Q \to \Semiring$, 
we shall use the auxiliary function $b_\top :  Q \times P \times Q \to \Phic$.
%
Intuitively, the function defined in \eqref{eq:btop}
associates to $c \in \Deltac$ 
the minimum weight of a computation of $A$
starting in state $q$ with a stack 
$\<c, p> \cdot \gamma \in \Gamma^+$ 
and ending in state $q'$ with the same stack,
such that the computation can not pop 
the pair made of $c$ and $p$ at the top of this stack,
but may only read these symbols.
Moreover, $A$ may push another pair $\< c', p'>$ %call symbols 
on the top of $\<c, p> \cdot \gamma$,
following the third case of 
in the definition \eqref{eq:SWVPA-weight} of $\weight_A$,
and may pop $\< c', p'>$ later, following the fifth case of \eqref{eq:SWVPA-weight} (return symbol). 
%However, it cannot apply one of the two last cases (return symbol and empty stack)
%when the current stack is $\gamma$.
%pop symbols in $\gamma$.
% Note that having a stack reduced to such a symbol makes impossible the application of the 
% two last cases in the definition of $\weight_A$ (return symbol and empty stack). 
% However, it is possible to apply the two first cases 
% (internal symbol or call symbol, with a push on the top of $\top$).

over~$\Delta$, $\Semiring$ and~$\bar\Phi$,
the minimal weight %\wrt~$\leq_\oplus$,
for a word in~$\Delta^*$.
%

\noindent
%More precisely,
We distinguish two cases : when the stack is empty, and when it is not.
\lydia{introduced 2 cases for b}
%
In the case of an empty stack, let $b_\bot : Q \times Q \to \Semiring$ be such that : %the function:
%
EQBBOT
%
The term $\config{q}{\bot}, s, \config{q'}{\bot}$
of this sum is the central expression in
the definition \eqref{eq:weightA} of $A(s_0)$, for the minimum $s_0$
of the function $\weight_A$.
\lydia{so ?}

%
\noindent
If the stack is not empty, let $\top$ be a fresh stack symbol which does not belong to $\Gamma$,
\florent{$b_\top$ : mot bien parenthèsé $c$/$r$}
and let $b_\top : Q \times P \times Q \to \Phic$ be such that,
for every two states $q, q' \in Q$
and stack symbol $p \in P$: %and $\sigma \in \{ \bot, \top \}$,
\begin{equation}\label{eq:btop}
  b_\top(q, p, q') : c \mapsto \bigoplus_{s\in \Delta^*}
  \textstyle
  \weight_A\bigl(\configup{q}{\< c, p> \stackup \top }, s, \configup{q'}{\<c, p> \stackup \top}\bigr)
\end{equation}
%

@@

\textbf{initially} let $\Q = (Q \times Q) \cup (Q \times P \times Q)$, %contains all
and let $d_\bot(q_1, q_2) = d_\top(q_1, p, q_2) = \one$
if $q_1 = q_2$ and $d_\bot(q_1, q_2) = d_\top(q_1, p, q_2) = \zero$ otherwise$\;$

@@

Algorithm~\ref{algo:Dijkstra}
constructs iteratively markings
$d_\bot : Q \times Q \to \Semiring$ and
$d_\top : Q \times P \times Q \to \Phic$
%of the triplets $\<q, \sigma, q'>$
%of states of $A$ by weight values in $\Semiring$,
that converges eventually to $b_\top$ and $b_\bot$.
%It uses for that purpose a priority queue $P$ containing triplets of
%$Q \times \{ \bot, \top \} \times Q$.

@@

\noindent
\florent{explication Fig.~\ref{fig:best-update} suivant cas de \eqref{eq:SWVPA-weight}}
The infinite sums in the updates of $d$ in Algorithm~\ref{algo:Dijkstra},
Figure~\ref{fig:best-update}
are well defined
since~$\Semiring$ is complete.
\florent{complete **}
** effectively computable by hypothesis that the label theory is effective**
The algorithm performs $2.|Q|^2$ iterations until $P$ is empty,
and each iteration has a time complexity $O(|Q|^2 . |P|)$.
\florent{detail with nb tr. and states}
That gives a time complexity $O(|Q|^4 . |P|)$.
It can be reduced by implementing $P$ as a priority queue,
prioritized by the value returned by $d$.
%$|Q|^3.\log(|Q|^2)$

The correctness of Algorithm~\ref{algo:Dijkstra}
is ensured by the invariant expressed in the following lemma.
\begin{lemma}\label{lem:bot}
For all $\< q_1, q_2> \notin \Q$,
$d_\bot(q_1, q_2) =  b_\bot(q_1, q_2)$/
\end{lemma}
The proof is by contradiction,
assuming a counter-example minimal in the length of the witness word.

\begin{lemma}\label{lem:top}
For all $\< q_1, p, q_2> \notin \Q$,
$d_\top(q_1, p, q_2) =  b_\top(q_1, p, q_2)$,
\end{lemma}

\noindent
For computing the minimal weight of a computation of $A$, we use the fact that,
at the termination of Algorithm~\ref{algo:Dijkstra}, %it holds that,
%There exist $q_1, q_2 \in Q$
\(
  {\displaystyle \bigoplus_{s \in \Delta^*} A(s)} =
  {\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
  \mathsf{in}(q) \mathop{\otimes} d_\bot(q, q') \mathop{\otimes} \mathsf{out}(q').
\)

\noindent
In order to obtain effectively a witness
(word of $\Delta^*$ with a computation of $A$ of minimal weight),
we require the additional property of convexity of weight functions.

@@

>>>>

\noindent
The weight associated by $A$ to $t \in \Delta^*$
is defined according to empty stack semantics:
%
\begin{equation}
A(t)  =
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr)
\mathop{\otimes} \mathsf{out}(q').
\label{eq:SWVPA-value}
\end{equation}

@@

%\subsection{Best Search}
\label{sec:best}\label{sec:search}
\noindent
We shall now present a procedure for searching, for a~$\SWVPA$ $A$,
a word %$w \in \Delta^*$
of minimal weight for~$A$,
as stated in the following proposition. %\wrt~$\leq_\oplus$,
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a Dijkstra algorithm computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
it holds that:
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
Since $\Semiring$ is complete, the infinite sum in~\eqref{eq:bbot} is well defined.
%
Following~\eqref{eq:SWVPA-value}, and the associativity and commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is:
\begin{equation}\label{eq:min}
{\displaystyle\bigoplus_{t\in \Delta^*}} A(t)
= 
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\end{equation}

\noindent
In order to compute the above function $b_\bot : Q \times Q \to \Semiring$, 
we shall consider an auxiliary function $b_\top :  Q \times P \times Q \to \Phic$.
%
Intuitively, when $b_\top(q, p, q') = \phi \in \Phic$, 
for all $c \in \Deltac$,  $\phi(c)$ is 
the minimum weight of a computation of~$A$
starting in state $q$ with a non-empty stack 
$\gamma' = \<c, p> \, \gamma \in \Gamma^+$ 
and ending in state $q'$ with the same stack $\gamma'$,
such that moreover, 
the computation does not pop the pair $\<c, p>$ at the top of $\gamma'$
(\ie $\gamma'$ is left untouched during the computation).
However, the computation can read $\<c, p>$ at the top of $\gamma'$,
and can also push another pair $\< c', p'> \in \Gamma$ on $\gamma'$,
following the third case of 
in the definition \eqref{eq:SWVPA-weight} of $\weight_A$ (call symbol).
The pair $\< c', p'>$ can be pop later, during the computation from $q$ to $q'$,  
following the fifth case of \eqref{eq:SWVPA-weight} (return symbol). 
%However, it cannot apply one of the two last cases (return symbol and empty stack)
%when the current stack is $\gamma$.
%pop symbols in $\gamma$.
%
Formally, in order to define $b_\top$, we consider 
a fresh stack symbol $\top \notin \Gamma$,   %which does not belong to $\Gamma$,
representing the above untouched stack, and let:
%
\begin{equation}\label{eq:btop}
  b_\top(q, p, q') : c \mapsto \bigoplus_{s\in \Delta^*}
  \textstyle
  \weight_A\bigl(\configup{q}{\< c, p> \stackup \top }, s, \configup{q'}{\<c, p> \stackup \top}\bigr)
\quad  
\mathrm{~for~all~} c \in \Deltac  
\end{equation}
%
By definition of $\weight_A$ in \eqref{eq:SWVPA-weight}, 
using the symbol $\top$ for the part of the stack below $\<c, p>$
(\ie the substack $\gamma$ in the above $\gamma' = \<c, p> \, \gamma$)
ensures that this part $\gamma'$ is not touched during the computation.
% there cannot be a pop of $\<c, p>$ followed by a push of $\<c, p>$
% because it is not possible to push over $\top$ (push read the top of stack).
%
This ensures in particular that the subword 
read during the computation is well parenthesized
(every symbol in $\Deltac$ has a matching symbol in $\Deltar$).

% Note that having a stack reduced to such a symbol makes impossible the application of the 
% two last cases in the definition of $\weight_A$ (return symbol and empty stack). 
% However, it is possible to apply the two first cases 
% (internal symbol or call symbol, with a push on the top of $\top$).

%The term $\config{q}{\bot}, s, \config{q'}{\bot}$
%of this sum is the central expression in
%the definition \eqref{eq:weightA} of $A(s_0)$, for the minimum $s_0$
%of the function $\weight_A$.

@@

\textbf{initially} let $\Q = (Q \times Q) \cup (Q \times P \times Q)$, 
and let $d_\bot(q_1, q_2) = d_\top(q_1, p, q_2) = \one$
if $q_1 = q_2$ and $d_\bot(q_1, q_2) = d_\top(q_1, p, q_2) = \zero$ otherwise$\;$

@@

Algorithm~\ref{algo:Dijkstra}
constructs iteratively two markings
$d_\bot : Q \times Q \to \Semiring$ and
$d_\top : Q \times P \times Q \to \Phic$,
%of the triplets $\<q, \sigma, q'>$
%of states of $A$ by weight values in $\Semiring$,
that converges eventually to $b_\top$ and $b_\bot$.
%It uses for that purpose a priority queue $P$ containing triplets of
%$Q \times \{ \bot, \top \} \times Q$.
%
\noindent
\florent{explication Fig.~\ref{fig:best-update} suivant cas de \eqref{eq:SWVPA-weight}}
The infinite sums in the updates of $d$ in Algorithm~\ref{algo:Dijkstra},
Figure~\ref{fig:best-update}
are well defined
since~$\Semiring$ is complete.

@@

\florent{complete **}
** effectively computable by hypothesis that the label theory is effective**
termination: The algorithm performs $2.|Q|^2$ iterations until $P$ is empty,
and each iteration has a time complexity $O(|Q|^2 . |P|)$.
\florent{detail with nb tr. and states}
That gives a time complexity $O(|Q|^4 . |P|)$.
It can be reduced by implementing $P$ as a priority queue,
prioritized by the value returned by $d$.
%$|Q|^3.\log(|Q|^2)$
%
At termination, $d_\bot = b_\bot$ and $d_\top = b_\top$
(see Appendix~\ref{sec:bestsearch}).
With~\eqref{eq:min}, this ensures
the correctness of Algorithm~\ref{algo:Dijkstra}.

\noindent
In order to obtain effectively a witness
(word of $\Delta^*$ with a computation of $A$ of minimal weight),
we require the additional property of convexity of weight functions.***
\florent{completer}

@@

\section{Correctness of the Best-Search Algorithm} \label{sec:bestsearch}
\input{app-bestsearch}
